```{r echo = F}

# Strings to match samples and create labels
comparison <- c("{{.x[1]}}", "{{.x[2]}}")

control <- comparison[1]
treatment <- comparison[2]

# Create section title
sect_title <- paste0(control, " vs ", treatment)


```

## **`r sect_title`**

```{r}
# Extract DE genes
sig_genes <- get_de(dds, params$sample_column, control, treatment,
                    write_csv = T, p_value = params$DE_alpha,
                    lfc = 0, output_dir = params$output_dir,
                    lfc_shrink = T)

sig_genes_high_lfc <- get_de(dds, params$sample_column, control, treatment,
                             write_csv = F, p_value = params$DE_alpha,
                             lfc = params$DE_lfc,
                             lfc_shrink = T)

DE_genes_n <- nrow(sig_genes$DE_genes)
```

### DEseq2 analysis
**`r format(DE_genes_n, big.mark = ",")`** differentially expressed genes were identified using the `DESeq2` package.

Differentially expressed genes are labeled on the MA plot in blue.
```{r, fig.width = 6, fig.height = 4}
plotMA(sig_genes$DE_df, main = paste0(control, " vs ", treatment))
```

### Heatmaps
Heatmap of DE genes across all samples
```{r, fig.width = 6, fig.height = 4}
heatmap_dir <- file.path(params$output_dir, "images", "heatmaps")
heatmap <- make_heatmap(dds = dds,
                        vsd = vsd,
                        de_genes = sig_genes$DE_genes,
                        treatment = treatment,
                        control = control,
                        group = params$sample_column,
                        print_genenames = FALSE,
                        cluster_cols = FALSE,
                        save_heatmap = TRUE,
                        output_dir = heatmap_dir,
                        color_test = sample_colors)
print(heatmap)

```

Heatmap of DE genes across only the samples used in the comparision

```{r, fig.width = 6, fig.height = 4}
heatmap_dir <- file.path(params$output_dir, "images", "heatmaps")
heatmap <- make_heatmap(dds = dds,
                        vsd = vsd,
                        de_genes = sig_genes$DE_genes,
                        treatment = treatment,
                        control = control,
                        group = params$sample_column,
                        print_genenames = FALSE,
                        cluster_cols = FALSE,
                        save_heatmap = FALSE,
                        plot_groups = c(treatment, control),
                        color_test = sample_colors)
print(heatmap)

```

### Gene Set Enrichment

For Gene Set Enrichment, I used `gprofiler2` It uses a hypergeometric test to determine overrepresentation of genes from different categories.

```{r, fig.height = 30, fig.width = 12}
all_gene_ens <- sub(".*_ENSMUSG", "ENSMUSG", rownames(dds))

result <- run_gprofiler_all(gene_table = sig_genes$DE_genes,
                            pos_name = control,
                            neg_name = treatment,
                            custom_bg = all_gene_ens,
                            save_dir = file.path(params$output_dir, "GSE_files"),
                            plot_dir = file.path(params$output_dir, "images",
                                                 "GSE"))

plots <- result$plots

plot_grid(plots$GOBP, plots$GOMF, plots$GOCC, plots$KEGG,
          nrow = 4,
          ncol = 1,
          align = "v",
          axis = "tb")

```